A Fast Unified Model for Parsing and Sentence Understanding

### 2016-05-29 writing/hard_stack_paper/paper.tex
Misc paper revisions.

In addition to the stack, buffer, and composition function, our full model includes an additional component: the tracking LSTM. This is a simple sequence-based LSTM RNN that operates in tandem with the model, taking inputs from the buffer and stack at each step. It is meant to maintain a low-resolution summary of the portion of the sentence that has been processed so far, which is used for two purposes: it supplies feature representations to the transition classifier, which allows the model to stand alone as a parser, and it additionally supplies a secondary input $\vec{e}$, see (\ref{eqn:lstm1}), to the composition function, allowing context information to enter the construction of sentence meaning,  forming what is effectively a tree-sequence hybrid model.

@@

For SPINN to operate on unparsed inputs, it needs to produce its own transition sequence $\mathbf a$ rather than relying on an external parser to supply it as part of the input. To do this, the model predicts $a_t$ at each step using a simple two-way softmax classifier whose input is the state of the tracking LSTM:
\begin{equation}
\vec{p}_{\text{a}} = \text{softmax}(W_{\text{trans}}\vec{h}_{\text{tracking}} + \vec{b}_{\text{trans}})
\end{equation}
Note that this is the close to being simplest viable implementation of a transition decision function. In contrast, the decision functions in sate-of-the-art transition-based parsers tend to use significantly richer feature sets as inputs, including features containing information about several upcoming words on the buffer instead of just one. We use this simple parser here, and leave a more thorough investigation of this component of the model to future work.

@@

\begin{table}[t]
\centering
\begin{tabular}{c @{\hspace*{2em}}c @{\hspace*{2em}} l}
  \toprule
  $t$ & $S$[$t$] & \multicolumn{1}{c}{$Q_t$} \\
  \midrule
  0 & & \underline{\hphantom{0} \hphantom{0}} \\
  1 & $a$ & \hphantom{0} \underline{\hphantom{0} 1} \\
  2 & $b$ & \hphantom{0} \hphantom{0} \underline{1 2} \\
  3 & $c$ & \hphantom{0} \hphantom{0} 1 \underline{2 3} \\
  4 & $(c~b)$ & \hphantom{0} \hphantom{0} \underline{1 4} \\
  5 & $((c~b)~a)$ & \hphantom{0} \underline{\hphantom{0} 5} \\
  \bottomrule
\end{tabular}
\caption{The thin-stack algorithm computing a \shift-\shift-\shift-\reduce-\reduce\ sequence on the input sentence $(a, b, c)$. $S$ is shown in the second column and represents the top of the stack at each step $t$. The last two elements of $Q$ (underlined) specify which rows $t$ would be involved in a \reduce\ operation at the next step.}
\label{tbl:thin-stack}
\end{table}

@@

We propose an alternative space-efficient stack representation inspired by the zipper technique \citep{huet1997zipper}, that we call \textit{thin stack}. For each input sentence, we represent the stack with a single $(T + 1) \times D$ matrix $S$. Each row $S_t$ represents the top of the actual stack at timestep $t$. At each timestep we can \shift\ a new element onto the stack, or \reduce\ the top two elements of the stack into a single element. To shift an element from the buffer to the top of the stack at timestep $t$, we simply write it into the location $S[t]$. In order to perform the \reduce\ operation, we need to retrieve the top two elements of the actual stack. We maintain a queue $Q$ of pointers into $S$ which contains the row indices of $S$ which are still present in the actual stack. The top two elements of the stack can be found by using the final two pointers in the queue $Q$. These retrieved rows are used to perform the \reduce\ operation, which modifies $Q$ to mark that some rows of $S$ have now been replaced in the actual stack. Algorithm~\ref{alg:thin-stack} describes the full mechanics of a stack feedforward in this compressed representation. It operates on the single $(T + 1) \times D$ matrix $S$ and a backpointer queue $Q$. Table~\ref{tbl:thin-stack} shows an example run.

@@

\vspace{0.5em}
{\noindent\small
{\bf Unlabeled binary parse:} ( ( the cat ) ( sat down ) )\\
{$\mathbf a$}: \shift, \shift, \reduce, \shift, \shift, \reduce, \reduce\\
{$\mathbf x$}: the, cat, sat, down
}

@@

In this section, we compare the test-time speed of our SPINN-PI-NT with an equivalent TreeRNN implemented in the conventional fashion and with a standard RNN sequence model. While the full models evaluated below are implemented and trained using Theano \citep{bergstra+al:2010-scipy,Bastien-Theano-2012}, which is reasonably efficient but not perfect for our model, we wish to compare well-optimized implementations of all three models. To do this, we reimplement the feedforward\footnote{We chose to reimplement and evaluate only the feedforward/inference pass, as inference speed is the relevant performance metric for most practical applications.} of SPINN-PI-NT and an LSTM RNN baseline in C++/CUDA, and compare that implementation with a CPU-based C++/Eigen TreeRNN implementation from \citet{irsoy2014deep}, which we modified to perform exactly the same computations as SPINN-PI-NT.\footnote{The original code for Irsoy \& Cardie's model is available at \url{https://github.com/oir/deep-recursive}. We plan to release our modified code at publication time, alongside the optimized C++/CUDA models and the Theano source code for the full SPINN.} TreeRNNs like this can only operate on a single example at a time and are thus poorly suited for GPU computation.

\begin{figure}
\centering
\resizebox{0.85\linewidth}{!}{
\begin{tikzpicture}
\begin{axis}[xmin=0, ymin=0, xlabel=Batch size, ylabel=Feedforward time (sec),
             xticklabels={0,32,64,128,256,512,1024,2048}, xmode=log, log basis x=2,
             legend style={at={(0.55, 0.5)},anchor=north west}]
  \addplot table [x={Batch size}, y=GPU, col sep=tab] {runtime.tsv};
  \addlegendentry{Thin-stack GPU}
  \addplot table [x={Batch size}, y=CPU, col sep=tab] {runtime.tsv};
  \addlegendentry{CPU \citep{irsoy2014deep}}
  \addplot table [x={Batch size}, y=RNN, col sep=tab] {runtime.tsv};
  \addlegendentry{RNN}
\end{axis}
\end{tikzpicture}
}
\caption{Feedforward speed comparison.}
\label{fig:speed}
\end{figure}

@@

To classify an SNLI sentence pair, we run two copies of SPINN with shared parameters: one on the premise sentence and another on the hypothesis sentence. We then use their outputs (the $\vec{h}$ states at the top of each stack at time $t=T$) to construct a feature vector $\vec{x}_{\text{classifier}}$ for the pair. This feature vector consists of the concatenation of these two sentence vectors, their difference, and their elementwise product \citep[following][]{mou2015recognizing}:
\begin{equation}
\vec{x}_{\text{classifier}} =
\colvec{4}
    {\vec{h}_{\text{premise}}}
    {\vec{h}_{\text{hypothesis}}}
    {\vec{h}_{\text{premise}} - \vec{h}_{\text{hypothesis}}}
    {\vec{h}_{\text{premise}} \odot \vec{h}_{\text{hypothesis}}}
\end{equation}
Following \citet{snli:emnlp2015}, this feature vector is then passed to a series of 1024D ReLU neural network layers (i.e., an MLP; the number of layers is tuned as a hyperparameter), then passed into a linear transformation, and then finally passed to a softmax layer, which yields a distribution over the three labels.

\paragraph{The objective function} Our objective combines a cross-entropy objective $\mathcal{L}^{\text{s}}$ for the SNLI classification task, a cross-entropy objective $\{\mathcal{L}_t^{\text{p}}, \mathcal{L}_t^{\text{h}}\}$ for each parsing decision for each of the sentences at each step $t$, and an L2 regularization term on the trained parameters. The terms are weighted using the tuned hyperparameters $\alpha$ and $\lambda$:
\begin{equation}
\begin{split}
\mathcal{L}_{\text{m}} = &\mathcal{L}^{\text{s}} + \alpha \sum_{t=0}^{T-1} (\mathcal{L}_t^{\text{p}} + \mathcal{L}_t^{\text{h}}) + \lambda \norm{\theta}^2_2
\end{split}
\end{equation}

@@

Finally, as is likely inevitable for models evaluated on SNLI, all four models under study did several percent worse on test examples whose ground truth label is \textit{neutral} than on examples of the other two classes. \textit{entailment}--\textit{neutral} and \textit{neutral}--\textit{contradiction} confusions appear to be much harder to avoid than \textit{entailment}--\textit{contradiction} confusions, where relatively superficial cues might be more readily useful.

>>>>

In addition to the stack, buffer, and composition function, our full model includes an additional component: the tracking LSTM. This is a simple sequence-based LSTM RNN that operates in tandem with the model, taking inputs from the buffer and stack at each step. It is meant to maintain a low-resolution summary of the portion of the sentence that has been processed so far, which is used for two purposes: it supplies feature representations to the transition classifier, which allows the model to stand alone as a parser, and it additionally supplies a secondary input $\vec{e}$ to the composition function---see (\ref{eqn:lstm1})---allowing context information to enter the construction of sentence meaning and forming what is effectively a tree-sequence hybrid model.

@@

For SPINN to operate on unparsed inputs, it needs to produce its own transition sequence $\mathbf a$ rather than relying on an external parser to supply it as part of the input. To do this, the model predicts $a_t$ at each step using a simple two-way softmax classifier whose input is the state of the tracking LSTM:
\begin{equation}
\vec{p}_{\text{a}} = \text{softmax}(W_{\text{trans}}\vec{h}_{\text{tracking}} + \vec{b}_{\text{trans}})
\end{equation}
This is the close to being simplest viable implementation of a transition decision function. In contrast, the decision functions in sate-of-the-art transition-based parsers tend to use significantly richer feature sets as inputs, including features containing information about several upcoming words on the buffer instead of just one supplied to the tracking LSTM.

@@

\begin{table}[t]
\centering
\begin{tabular}{cc @{\hspace*{2em}}c @{\hspace*{2em}} l}
  \toprule
  $t$ & $a_t$ & $S$[$t$] & \multicolumn{1}{c}{$Q_t$} \\
  \midrule
  0 & \shift &  & \underline{\hphantom{0} \hphantom{0}} \\
  1 & \shift & \word{Spot} & \hphantom{0} \underline{\hphantom{0} 1} \\
  2 & \shift & \word{sat} & \hphantom{0} \hphantom{0} \underline{1 2} \\
  3 & \reduce & \word{down} & \hphantom{0} \hphantom{0} 1 \underline{2 3} \\
  4 & \reduce & \word{sat down} & \hphantom{0} \hphantom{0} \underline{1 4} \\
  5 & & \word{Spot} (\word{sat down}) & \hphantom{0} \underline{\hphantom{0} 5} \\
  \bottomrule
\end{tabular}
\caption{The thin-stack algorithm operating on the input sequence $\mathbf x = \text{(\word{Spot}, \word{sat}, \word{down})}$ and the transition sequence shown in the second column. $S$ is shown in the third column and represents the top of the stack at each step $t$. The last two elements of $Q$ (underlined) specify which rows $t$ would be involved in a \reduce\ operation at the next step.}
\label{tbl:thin-stack}
\end{table}

@@

We propose an alternative space-efficient stack representation inspired by the zipper technique \citep{huet1997zipper}, that we call \textit{thin stack}. For each input sentence, we represent the stack with a single $T \times D$ matrix $S$. Each row $S[t]$ (for $0<t\le T$) represents the top of the actual stack at timestep $t$. At each timestep we can \shift\ a new element onto the stack, or \reduce\ the top two elements of the stack into a single element. To shift an element from the buffer to the top of the stack at timestep $t$, we simply write it into the location $S[t]$. In order to perform the \reduce\ operation, we need to retrieve the top two elements of the actual stack. We maintain a queue $Q$ of pointers into $S$ which contains the row indices of $S$ which are still present in the actual stack. The top two elements of the stack can be found by using the final two pointers in the queue $Q$. These retrieved rows are used to perform the \reduce\ operation, which modifies $Q$ to mark that some rows of $S$ have now been replaced in the actual stack. Algorithm~\ref{alg:thin-stack} describes the full mechanics of a stack feedforward in this compressed representation. It operates on the single $T \times D$ matrix $S$ and a backpointer queue $Q$. Table~\ref{tbl:thin-stack} shows an example run.

@@

\vspace{0.5em}
{\noindent\small
{\bf Unlabeled binary parse:} ( ( the cat ) ( sat down ) )\\
{$\mathbf x$}: \word{the}, \word{cat}, \word{sat}, \word{down}\\
{$\mathbf a$}: \shift, \shift, \reduce, \shift, \shift, \reduce, \reduce
}

@@

In this section, we compare the test-time speed of our SPINN-PI-NT with an equivalent TreeRNN implemented in the conventional fashion and with a standard RNN sequence model. While the full models evaluated below are implemented and trained using Theano \citep{theano}, which is reasonably efficient but not perfect for our model, we wish to compare well-optimized implementations of all three models. To do this, we reimplement the feedforward\footnote{We chose to reimplement and evaluate only the feedforward/inference pass, as inference speed is the relevant performance metric for most practical applications.} of SPINN-PI-NT and an LSTM RNN baseline in C++/CUDA, and compare that implementation with a CPU-based C++/Eigen TreeRNN implementation from \citet{irsoy2014deep}, which we modified to perform exactly the same computations as SPINN-PI-NT.\footnote{The original code for Irsoy \& Cardie's model is available at \url{https://github.com/oir/deep-recursive}. Our optimized C++/CUDA models and the Theano source code for the full SPINN are available at \url{https://github.com/hans/spinn}.} TreeRNNs like this can only operate on a single example at a time and are thus poorly suited for GPU computation.

\begin{figure}
\centering
\resizebox{1\linewidth}{!}{
\begin{tikzpicture}
\begin{axis}[xmin=0, ymin=0, xlabel=Batch size, ylabel=Feedforward time (sec),
             xticklabels={0,32,64,128,256,512,1024,2048}, xmode=log, log basis x=2,
             legend style={at={(0.55, 0.5)},anchor=north west}]
  \addplot table [x={Batch size}, y=GPU, col sep=tab] {runtime.tsv};
  \addlegendentry{Thin-stack GPU}
  \addplot table [x={Batch size}, y=CPU, col sep=tab] {runtime.tsv};
  \addlegendentry{CPU \citep{irsoy2014deep}}
  \addplot table [x={Batch size}, y=RNN, col sep=tab] {runtime.tsv};
  \addlegendentry{RNN}
\end{axis}
\end{tikzpicture}
}
\caption{Feedforward speed comparison.}
\label{fig:speed}
\end{figure}

@@

To classify an SNLI sentence pair, we run two copies of SPINN with shared parameters: one on the premise sentence and another on the hypothesis sentence. We then use their outputs (the $\vec{h}$ states at the top of each stack at time $t=T$) to construct a feature vector $\vec{x}_{\text{classifier}}$ for the pair. This feature vector consists of the concatenation of these two sentence vectors, their difference, and their elementwise product \citep[following][]{mou2015recognizing}:
\begin{equation}
\vec{x}_{\text{classifier}} =
\colvec{4}
    {\vec{h}_{\text{premise}}}
    {\vec{h}_{\text{hypothesis}}}
    {\vec{h}_{\text{premise}} - \vec{h}_{\text{hypothesis}}}
    {\vec{h}_{\text{premise}} \odot \vec{h}_{\text{hypothesis}}}
\end{equation}
This feature vector is then passed to a series of 1024D ReLU neural network layers (i.e., an MLP; the number of layers is tuned as a hyperparameter), then passed into a linear transformation, and then finally passed to a softmax layer, which yields a distribution over the three labels.

\paragraph{The objective function} Our objective combines a cross-entropy objective $\mathcal{L}_{\text{s}}$ for the SNLI classification task, cross-entropy objectives $\mathcal{L}^t_{\text{p}}$ and $\mathcal{L}^t_{\text{h}}$ for the parsing decision for each of the two sentences at each step $t$, and an L2 regularization term on the trained parameters. The terms are weighted using the tuned hyperparameters $\alpha$ and $\lambda$:
\begin{equation}
\begin{split}
\mathcal{L}_{\text{m}} = &\mathcal{L}_{\text{s}} + \alpha \sum_{t=0}^{T-1} (\mathcal{L}^t_{\text{p}} + \mathcal{L}^t_{\text{h}}) + \lambda \norm{\theta}^2_2
\end{split}
\end{equation}

@@

Finally, as is likely inevitable for models evaluated on SNLI, all four models under study did several percent worse on test examples whose ground truth label is \textit{neutral} than on examples of the other two classes. \textit{Entailment}--\textit{neutral} and \textit{neutral}--\textit{contradiction} confusions appear to be much harder to avoid than \textit{entailment}--\textit{contradiction} confusions, where relatively superficial cues might be more readily useful.
